\pkg{cubble}: An R Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2022-06-24 4-1-basic.Rmd
fix citation on websites

First, the stations need to be filtered to those in NSW and Victoria by using the station identifiers, stored within the 11 digits of the \code{id} variable entries. The country code is in the first 5 digits (Australia is represented by "ASN00") and the next 6 digits encode the station following the [Australian Bureau of Meteorology (BOM)](http://www.bom.gov.au/climate/cdo/about/site-num.shtml) coding protocols. New South Wales stations correspond to entries in the range 46-75 and the Victorian stations to 76-90. Filtering Victoria and New South Wales stations is a *spatial operation* and hence uses the \code{cubble} nested form:

>>>>

First, the stations need to be filtered to those in NSW and Victoria by using the station identifiers, stored within the 11 digits of the \code{id} variable entries. The country code is in the first 5 digits (Australia is represented by "ASN00") and the next 6 digits encode the station following the Australian Bureau of Meteorology (BOM) [@bom] coding protocols. New South Wales stations correspond to entries in the range 46-75 and the Victorian stations to 76-90. Filtering Victoria and New South Wales stations is a *spatial operation* and hence uses the \code{cubble} nested form:

<!-- BOM link: http://www.bom.gov.au/climate/cdo/about/site-num.shtml -->

### 2022-06-24 4-3-water-level.Rmd
fix citation on websites

River level and rainfall data for the same areas should have some similarity. Here we examine the river gauge data (\code{Water_course_level}) from the Bureau of Meteorology [@REF] in relation to weather station rainfall from NOAA's climate data [@REF] (\code{climate}). Figure \ref{fig:matching-map} shows the location of available weather stations and water gauges in Victoria. The goal is to match water gauges with nearby weather stations, spatially and temporally, using 2020 measurements using 
\code{match_sites()} function.
<!--The Bureau of Meteorology collects [water data](http://www.bom.gov.au/metadata/catalogue/19115/ANZCW0503900528?template=full) from river gauges. The collected variables include: electrical conductivity, turbidity, watercourse discharge, watercourse level, and water temperature. We expect that the water level should be related to the precipitation available from climate data, since rainfall will raise the water level in the river. Figure \ref{fig:matching-map} shows the location of available weather stations and water gauges in Victoria, and we will try to match water gauges with nearby weather stations, taking into account both spatial and temporal information from 2020.-->

@@

This function requires passing the major and minor data sets used for matching, in this case those are \code{river} and \code{climate}.  The variables used for the temporal matching  are \code{Water_course_level} from the \code{river} data set and \code{prcp} in the climate data set. The rest of the arguments, as explained in Section \@ref(sec:matching), correspond to the maximum and minimum number of peaks in the time series to be matched. In this example those are set to be a maximum of 30 and a minimum of 15 (approximately 2 and 1 per month).

>>>>

River level and rainfall data for the same areas should have some similarity. Here we examine the river gauge data (\code{Water_course_level}) from the Bureau of Meteorology [@bom] in relation to weather station rainfall from NOAA's climate data [@noaa] (\code{climate}). Figure \ref{fig:matching-map} shows the location of available weather stations and water gauges in Victoria. The goal is to match water gauges with nearby weather stations, spatially and temporally, using 2020 measurements using 
\code{match_sites()} function.
<!--The Bureau of Meteorology collects [water data](http://www.bom.gov.au/metadata/catalogue/19115/ANZCW0503900528?template=full) from river gauges. The collected variables include: electrical conductivity, turbidity, watercourse discharge, watercourse level, and water temperature. We expect that the water level should be related to the precipitation available from climate data, since rainfall will raise the water level in the river. Figure \ref{fig:matching-map} shows the location of available weather stations and water gauges in Victoria, and we will try to match water gauges with nearby weather stations, taking into account both spatial and temporal information from 2020.-->

@@

This function requires passing the major and minor data sets used for matching, in this case those are \code{river} and \code{climate}.  The variables used for the temporal matching  are \code{Water_course_level} from the \code{river} data set and \code{prcp} in the climate data set. The rest of the arguments, as explained in Section \@ref(matching), correspond to the maximum and minimum number of peaks in the time series to be matched. In this example those are set to be a maximum of 30 and a minimum of 15 (approximately 2 and 1 per month).

### 2022-06-24 4-4-netcdf.Rmd
fix citation on websites

Figure \ref{fig:netcdf} reproduces the ERA5 data row of Figure 19 in @hersbach2020era5. Here we explain how this would be done using cubble. The plots show that the southern polar vortex splits into two on 2002-09-26 and further splits into four on 2002-10-04. Further explanation of why this is interesting can be found in the figure source, and also @simmons2020global, @simmons2005ecmwf.

The ERA5 data [@hersbach2020era5] provides hourly estimates across the Earth for atmospheric, land and oceanic climate variables. The data is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF). It can be directly downloaded from [Copernicus Climate Data Store (CDS)](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels?tab=overview) website or via the R package \pkg{ecmwfr} [@ecwmfr].

>>>>

Figure \ref{fig:netcdf} reproduces the ERA5 data [@hersbach2020era5] row of Figure 19 in [@hersbach2020era5]. Here we explain how this would be done using cubble. The plots show that the southern polar vortex splits into two on 2002-09-26 and further splits into four on 2002-10-04. Further explanation of why this is interesting can be found in the figure source, and also [@simmons2020global], [@simmons2005ecmwf].

The ERA5 data [@hersbach2020era5] provides hourly estimates across the Earth for atmospheric, land and oceanic climate variables. The data is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF). It can be directly downloaded from Copernicus Climate Data Store (CDS) [@cds] website or via the R package \pkg{ecmwfr} [@ecwmfr].

### 2022-06-24 5-conclusion.Rmd
fix citation on websites

There are several possible future directions for the work. The data structure only described fixed spatial sites, and it could be useful to provide tools to accommodate moving coordinates as might be encountered in animal movement data. That could be achieved with a list-column for the location coordinates, and an additional form that these locations can be pivoted into, like the long form for temporal variables. For multiple measured variables, \pkg{cubble} could be integrated with dimension reduction methods, and dynamic graphics for multiple dimensions such as a tour [@REF].

>>>>

There are several possible future directions for the work. The data structure only described fixed spatial sites, and it could be useful to provide tools to accommodate moving coordinates as might be encountered in animal movement data. That could be achieved with a list-column for the location coordinates, and an additional form that these locations can be pivoted into, like the long form for temporal variables. For multiple measured variables, \pkg{cubble} could be integrated with dimension reduction methods, and dynamic graphics for multiple dimensions such as a tour [@wickham_tourr_2011].

### 2022-06-24 jss/cubble-jss.tex
fix citation on websites

First, the stations need to be filtered to those in NSW and Victoria by using the station identifiers, stored within the 11 digits of the \code{id} variable entries. The country code is in the first 5 digits (Australia is represented by ``ASN00'') and the next 6 digits encode the station following the \href{http://www.bom.gov.au/climate/cdo/about/site-num.shtml}{Australian Bureau of Meteorology (BOM)} coding protocols. New South Wales stations correspond to entries in the range 46-75 and the Victorian stations to 76-90. Filtering Victoria and New South Wales stations is a \emph{spatial operation} and hence uses the \code{cubble} nested form:

@@

River level and rainfall data for the same areas should have some similarity. Here we examine the river gauge data (\code{Water_course_level}) from the Bureau of Meteorology \citep{REF} in relation to weather station rainfall from NOAA's climate data \citep{REF} (\code{climate}). Figure \ref{fig:matching-map} shows the location of available weather stations and water gauges in Victoria. The goal is to match water gauges with nearby weather stations, spatially and temporally, using 2020 measurements using
\code{match_sites()} function.

@@

This function requires passing the major and minor data sets used for matching, in this case those are \code{river} and \code{climate}. The variables used for the temporal matching are \code{Water_course_level} from the \code{river} data set and \code{prcp} in the climate data set. The rest of the arguments, as explained in Section \ref{sec:matching}, correspond to the maximum and minimum number of peaks in the time series to be matched. In this example those are set to be a maximum of 30 and a minimum of 15 (approximately 2 and 1 per month).

@@

Figure \ref{fig:netcdf} reproduces the ERA5 data row of Figure 19 in \citet{hersbach2020era5}. Here we explain how this would be done using cubble. The plots show that the southern polar vortex splits into two on 2002-09-26 and further splits into four on 2002-10-04. Further explanation of why this is interesting can be found in the figure source, and also \citet{simmons2020global}, \citet{simmons2005ecmwf}.

The ERA5 data \citep{hersbach2020era5} provides hourly estimates across the Earth for atmospheric, land and oceanic climate variables. The data is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF). It can be directly downloaded from \href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels?tab=overview}{Copernicus Climate Data Store (CDS)} website or via the R package \pkg{ecmwfr} \citep{ecwmfr}.

@@

There are several possible future directions for the work. The data structure only described fixed spatial sites, and it could be useful to provide tools to accommodate moving coordinates as might be encountered in animal movement data. That could be achieved with a list-column for the location coordinates, and an additional form that these locations can be pivoted into, like the long form for temporal variables. For multiple measured variables, \pkg{cubble} could be integrated with dimension reduction methods, and dynamic graphics for multiple dimensions such as a tour \citep{REF}.

>>>>

First, the stations need to be filtered to those in NSW and Victoria by using the station identifiers, stored within the 11 digits of the \code{id} variable entries. The country code is in the first 5 digits (Australia is represented by ``ASN00'') and the next 6 digits encode the station following the Australian Bureau of Meteorology (BOM) \citep{bom} coding protocols. New South Wales stations correspond to entries in the range 46-75 and the Victorian stations to 76-90. Filtering Victoria and New South Wales stations is a \emph{spatial operation} and hence uses the \code{cubble} nested form:

@@

River level and rainfall data for the same areas should have some similarity. Here we examine the river gauge data (\code{Water_course_level}) from the Bureau of Meteorology \citep{bom} in relation to weather station rainfall from NOAA's climate data \citep{noaa} (\code{climate}). Figure \ref{fig:matching-map} shows the location of available weather stations and water gauges in Victoria. The goal is to match water gauges with nearby weather stations, spatially and temporally, using 2020 measurements using
\code{match_sites()} function.

@@

This function requires passing the major and minor data sets used for matching, in this case those are \code{river} and \code{climate}. The variables used for the temporal matching are \code{Water_course_level} from the \code{river} data set and \code{prcp} in the climate data set. The rest of the arguments, as explained in Section \ref{matching}, correspond to the maximum and minimum number of peaks in the time series to be matched. In this example those are set to be a maximum of 30 and a minimum of 15 (approximately 2 and 1 per month).

@@

Figure \ref{fig:netcdf} reproduces the ERA5 data \citep{hersbach2020era5} row of Figure 19 in \citep{hersbach2020era5}. Here we explain how this would be done using cubble. The plots show that the southern polar vortex splits into two on 2002-09-26 and further splits into four on 2002-10-04. Further explanation of why this is interesting can be found in the figure source, and also \citep{simmons2020global}, \citep{simmons2005ecmwf}.

The ERA5 data \citep{hersbach2020era5} provides hourly estimates across the Earth for atmospheric, land and oceanic climate variables. The data is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF). It can be directly downloaded from Copernicus Climate Data Store (CDS) \citep{cds} website or via the R package \pkg{ecmwfr} \citep{ecwmfr}.

@@

There are several possible future directions for the work. The data structure only described fixed spatial sites, and it could be useful to provide tools to accommodate moving coordinates as might be encountered in animal movement data. That could be achieved with a list-column for the location coordinates, and an additional form that these locations can be pivoted into, like the long form for temporal variables. For multiple measured variables, \pkg{cubble} could be integrated with dimension reduction methods, and dynamic graphics for multiple dimensions such as a tour \citep{wickham_tourr_2011}.
